---
title: "Prediction of Weight Lifting Quality"
author: "AT"
date: "Saturday, September 26, 2015"
output: html_document
---

### I. Executive Summary
In this study, we attempt to predict the quality in which six male participants performed Unilateral Dumbbell Biceps Curls, by relying on the measurements of the actions of the participants who wear sensors located on the arm, forearm, belt and dumbbells. 


### II. Background
Six male participants aged from 20 to 28 years, with little weight lifting experience, were asked to perform 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different ways: (A) according to specification, (B) throwing the elbow to the front, (C) lifting the dumbbells halfway, (D) lowering the dumbbells only halfway and (E) throwing the hips to the front. Besides A, the other four methods were incorrect ways of weight lifting.

Measurements were taken from the sensors that participants wore on their arms, forearms, belt and the dumbbells they used. Since the participants were relatively inexperienced at weight lifting, they were monitored by an experienced weight lifter and also used a somewhat light dumbbell weighing 1.25 kg.

### III. Treating the Data
Our training data was obtained from [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv). 

```{r, cache=TRUE}
dat <- read.csv("pml-training.csv")
```

There are `r nrow(dat)` observations and `r length(dat)` variables, of which the "__classe__" variable tells us whether the observation is correct weight lifting or not.

To facilitate cross validation later on, I have made use of the **K-Fold** method in the **caret** package to slice the data into 3 parts, where I will use the 2/3 part to build a model and the remaining 1/3 for cross validation.


```{r}
require(caret)
Ind <- createFolds(y=dat$classe, k=3, list=T)
training <- dat[-Ind[[1]],]
valid <- dat[Ind[[1]],]

```

Many of the variables contain $NA$ or are empty, hence I will not be choosing them for the model building. 

Here are the column numbers for the variables I will zoom in on, the last one being the classe variable.

```{r}
var_col <- c(8:10, 37:48, 60:68, 84:86, 113:124, 151:160)
training1 <- training[, var_col]
valid1 <- valid[, var_col]
any(is.na(training1))
any(is.na(valid1))
```

We can see that there are no NA values in the new dataset.


### IV. Training the Data

Because there are 48 variables, which is a considerably large number of variables, we first try to narrow down this number by using principal component analysis. 

```{r}
library(gbm)
preProc <- preProcess(training1[,-49],method="pca", thresh=0.95)
```

Following that, we applied the generalised boosted regression method to model this data.

```{r, cache=TRUE}
trainPC <- predict(preProc,training1[,-49])
modelFit <- train(training1$classe ~ ., method="gbm", data=trainPC, verbose=F)
modelFit
```

We then test the model on the the validation dataset, which was borne out of our original training set.

```{r}
validatePC <- predict(preProc,valid1[,-49])
confusionMatrix(valid1$classe, predict(modelFit,validatePC))
```

The accuracy was found to be quite high, around 81%. The out of sample error is estimated to be 18.67%.


### V. Predicting 

Finally, we attempt to use the model to predict the test sample, which can be downloaded [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv). This test dataset contains 20 observations.

We read the test data from our working directory, assign it the values for the variable identified by principal component analysis.

```{r}
testdata <- read.csv("pml-testing.csv")
testdata <- testdata[,var_col]
testPC <- predict(preProc, testdata[,-49])
```

We then predict the class variable that each of the 20 observations belong to.

```{r}
predict(modelFit, newdata=testPC)
```


### VI. References

1. Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

(Read more: <http://groupware.les.inf.puc-rio.br/har#ixzz3msvM066o>)

